---
title: "Chapter 6 Summary Statistics For One Variable"
output: github_document
---

```{r}
library(gapminder)
library(tidyverse)
```

## 6.1 One Quantitative Variable

### 6.1.1 Using base R summary function

```{r}
gapminder %>% filter(year==1997) %>% select(lifeExp) %>% summary()
```

### 6.1.2 Using dplyr summarise function

```{r}
gapminder %>% dplyr::filter(year==1997) %>% 
      dplyr::summarise(meanLE=mean(lifeExp,na.rm=TRUE),
                       medLE=median(lifeExp,na.rm=TRUE),
                       sd=sd(lifeExp,na.rm=TRUE),
                       iqr=IQR(lifeExp,na.rm=TRUE),
                      Q1=quantile(lifeExp,probs=0.25,na.rm=TRUE),
                      Q3=quantile(lifeExp,probs=0.75),
                      n=n())
```

### 6.1.3 Summary Statistics Using funModeling package

```{r}
funModeling::profiling_num(gapminder)

```

### 6.1.4 Summary Statistics: skimr package

The *skimr* package produces summary statistics about variables and overviews for dataframes. It is easy to manipulate and use pipes, select, and filter from the *tidyverse* family of packages.

The next code supplies a dataframe that contains both categorical variables (continent), and numeric variables (lifeExp, gdpPercap). Numeric variables are chosen with the `yank` function, then some attributes are omitted from the display (n_missing, complete_rate) using the `select` function from dplyr.

```{r}
varlist <- c("n_missing","complete_rate")
gapminder %>% filter(year==1997) %>% 
  select(-year, -country, -pop) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("numeric") %>%
  dplyr::select(-one_of(varlist))
```

## 6.2 One Categorical Variable

### 6.2.1 Counting Values

```{r}
gapminder %>% count(continent)
```

### 6.2.2 Categorical variable: skimr package

```{r}
gapminder %>% filter(year==1997) %>% 
  select(lifeExp,continent) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("factor") %>%
  dplyr::select(-n_missing,-ordered,-complete_rate)
```

### 6.2.3 Categorical variable: funModeling package

The *funModeling* package gives an easy way to learn about categorical variables of types: character and factor. There are two categorical variables in the gapminder dataframe: country and continent. There are a lot of countries, so we demonstrate this command for only the continent variable.

```{r}
# Frequency distribution of entire dataframe
# will produce lots of output and warnings
#funModeling::freq(gapminder)
# next command for one category variable:  continent
funModeling::freq(gapminder$continent)
```

### 6.2.4 Categorical variable: janitor package

```{r}
gapminder %>% 
  filter(year==1997) %>%
  select(continent) %>%
  table()
```

```{r}
#
gapminder %>% 
  filter(year==1997) %>%
  janitor::tabyl(continent,sort=TRUE) %>%
  janitor::adorn_pct_formatting(digits=2,affix_sign = TRUE) %>%
  knitr::kable()
```
