---
title: "Chapter 3 Overview of a Dataframe"
output: github_document
---

```{r}
library(gapminder)
library(tidyverse)
```

## 3.1 glimpse

Let's look at what is inside the gapminder dataset using the `glimpse` command from the *dplyr* package. The *dplyr* package is contained in the package "tidyverse" that was loaded previously. The `glimpse(gapminder)` command would have executed without any errors. We use the `dplyr::` prefix to inform readers that the glimpse function resides in the *dplyr* package.

```{r}
dplyr::glimpse(gapminder)
```

Each variable name is listed along with a variable type designation.

-   fct: means a factor variable, also known as a categorical variable.

-   int: means a quantitative variable that takes only integer or whole number values.

-   dbl: means double precision, a quantitative variable that is essentially continuous - taking decimal values.

## 3.2 head

By default, the `head` command will show the first 6 rows of the dataset gapminder. Datasets in R are called "dataframes." The gapminder dataframe is denoted as a "tibble" which is a type of dataframe.

Options to the `head` command can change the rows displayed.

```{r}
# default is to show 6 rows
#head(gapminder)
head(gapminder,n=13)
```

## 3.3 tail

```{r}
tail(gapminder)
```

## 3.4 summary

```{r}
summary(gapminder)
```

```{r}
# usage of pipe operator
# Same idea, but using tidyverse pipe
gapminder %>% summary()

```

## 3.5 Dataframe Details: funModeling package

The *funModeling* package contains the `df_status` command which also summarizes a dataframe - showing different aspects like missing values, percentage of zero values, and also the number of unique values.

```{r}
funModeling::df_status(gapminder)
```

```{r}
funModeling::data_integrity(gapminder)
```

## 3.6 Dataframe Details: skimr package

```{r}
gapminder %>% 
  skimr::skim_without_charts() 

```

## 3.7 describe: Hmisc package

```{r}
gapminder %>% 
  Hmisc::describe() 
```

```{r}
```

```{r}

```

```{r}

```
