---
title: "Chapter 7 Exploratory Data Analysis For One Quantitative Variable: by Groups & Chapter 8 Analysis of One Categorical Variable by another categorical variable"
output: github_document
---

```{r}
library(gapminder)
library(tidyverse)
```

## 7.1 Summary Statistics: dplyr

```{r}
# Output presented in initial continent order (alphabetic)
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            Q1=quantile(lifeExp, probs=0.25,na.rm=TRUE),
            Q3=quantile(lifeExp,probs=0.75),
            n=n())
```

```{r}
#
# Output rows ordered by decreasing values of a statistic (mean Life Expectancy):
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            min=min(lifeExp),
            max=max(lifeExp),
            n=n())  %>%
  arrange(desc(meanLE))
```

```{r}
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  summarise(meanLE=mean(lifeExp,na.rm=TRUE),
            medLE=median(lifeExp,na.rm=TRUE),
            sd=sd(lifeExp,na.rm=TRUE),
            iqr=IQR(lifeExp,na.rm=TRUE),
            min=min(lifeExp),
            max=max(lifeExp),
            n=n(),.groups = 'drop')  %>% 
  arrange(desc(meanLE)) %>%
  knitr::kable()
```

## 7.2 Summary Statistics: skimr

```{r}
gapminder %>% filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>% 
  skimr::skim_without_charts() %>%
  skimr::yank("numeric") %>%
  dplyr::filter(skim_variable %in% c("lifeExp","gdpPercap")) %>%
  knitr::kable()
```

## 7.3 Graphical Displays of a quantitative variable, separated by groups

In each example, the first lines create the dataset to be graphed - followed by a `ggplot` command making the display. Several of the examples make use of the principle of "small-multiples" so that each level of the factor variable has a separarate panel for the quantitative variable display.

### 7.3.1 Dotplots

```{r}
ds <- gapminder %>% filter(year==1997) 
#
ggplot(data=ds,mapping=aes(x=lifeExp)) + 
  geom_dotplot() + 
  facet_wrap( ~ continent,ncol=2) + 
  xlab("Life Expectancy (years)") +
  ylab("Frequency")
```

### 7.3.2 Histograms

```{r}
gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
ggplot(data=., mapping=aes(x=lifeExp)) + 
  geom_histogram(binwidth=5) + 
  facet_wrap( ~ continent,ncol=2) + 
  xlab("Life Expectancy (years)") +
  ylab("Frequency")
```

### 7.3.3 Density Plots in Facets

```{r}
gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
ggplot(data=., mapping=aes(x=lifeExp, colour=continent, fill=continent)) + 
  geom_density(alpha = 0.35) + 
  xlab("Life Expectancy (years)") +
  ylab("Density") +
  facet_wrap( ~ continent, ncol = 2) +  
  theme(legend.position = "none")
```

### 7.3.4 Overlaid Density Plots

```{r}
gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) %>%
ggplot(data=., mapping=aes(x=lifeExp, colour=continent, fill=continent)) + 
  geom_density(alpha = 0.35) + 
  xlab("Life Expectancy (years)") +
  ylab("Density")
```

### 7.3.5 Boxplots, Grouped Data

```{r}
ds <- gapminder %>% 
  filter(year==1997) %>% 
  filter(continent != "Oceania") %>% 
  group_by(continent) 
#
ggplot(data=ds, mapping=aes(x=continent,y=lifeExp)) +
 geom_boxplot() + 
  labs(x="Continent",y="Life Expectancy (years)")
```

## 8.1 Tables

```{r}
congage <- fivethirtyeight::congress_age
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) 
#  We declare party and chamber as factor/categorical variables, and control their levels.
ds1 <- ds1 %>% mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber))
ds1 <- ds1 %>% na.omit()
ds <- ds1  
#
table(ds$chamber,ds$party)
```

```{r}
#
mytable <- table(ds$chamber,ds$party)
#
prop.table(mytable) # cell percentages
```

```{r}
# Use package janitor
ds %>%  janitor::tabyl(chamber, party)
```

```{r}
# row percentages
t2 <- ds %>%  janitor::tabyl(chamber, party)
t2 %>%
  janitor::adorn_percentages("row") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
```

```{r}
# column percentages
t2 %>%
  janitor::adorn_percentages("col") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
```

```{r}
# both row and column percentages
t2 %>%
  janitor::adorn_percentages("all") %>%
  janitor::adorn_pct_formatting(digits = 2) %>%
  janitor::adorn_ns()
```

```{r}
congage <- fivethirtyeight::congress_age
ds1 <- congage %>% filter(congress > 112) %>% select(congress,chamber,state,party,incumbent,age) 
#  We declare party and chamber as factor/categorical variables, and control their levels.
ds1 <- ds1 %>% mutate(party = factor(party,levels=c("D","I","R")),
                    chamber = factor(chamber))
ds1 <- ds1 %>% na.omit()
ds <- ds1  
ds %>% group_by(chamber,party) %>% 
  dplyr::count() %>% 
  tidyr::pivot_wider(names_from = party, values_from = n)
```

## 8.2 Graphical Displays

```{r}
# basic bar plot of party affiliation
ggplot(data=ds, mapping=aes(x=party)) + 
  geom_bar() +
  labs(x="Party", y="Count")
```

```{r}
ds <- ds1 %>% group_by(party,chamber)
#
ggplot(data=ds, aes(x=chamber)) + 
  geom_bar(aes(fill=party),position="dodge") +
  labs(x="Chamber", y="Count")
```

```{r}
#
ggplot(data=ds, aes(x=party)) + 
  geom_bar(aes(fill=party)) +
  facet_wrap( ~ chamber) +
  labs(x="Party", y="Count")
```
