---
title: "Chapter 4 Introduction to Data Wrangling"
output: github_document
---

```{r}
library(gapminder)
library(tidyverse)
```

## 4.1 Tidy Data

What is tidy data? Tidy data is a standard way of mapping the meaning of a dataset to its structure. A dataset is messy or tidy depending on how rows, columns and tables are matched up with observations, variables and types. In tidy data:

-   Each variable forms a column.

-   Each observation forms a row.

-   Each type of observational unit forms a table.

## 4.2 Subset using filter

```{r}
#gapminder %>% filter(country=="Australia") %>% head(n=12)
gapminder %>% dplyr::filter(country=="Australia") %>% head(n=12)
```

## 4.3 Subset using multiple conditions

```{r}
gapminder %>% 
  dplyr::filter(continent=="Oceania" & year==1997) %>% 
  head()

```

## 4.4 Saving as a new dataframe

Here we save the the modified dataset as a new dataframe called gap97.

```{r}
gap97 <- gapminder %>% 
  filter(continent!="Oceania" & year==1997) 
#
dplyr::glimpse(gap97)
```

```{r}
gap97%>%summary()
```

## 4.5 Subset using select

```{r}
# the next command selects three variables and renames two of them:
gapminder %>% 
  dplyr::select(country, Year=year,LifeExp=lifeExp) %>% 
  head(n=12)
```

The `profiling_num` command from the *funModeling* package produces a lot of output, some we might not want. We will show how to modify the output of this command here. The command produces a dataframe which has many columns we might not wish to display or consider further.

We begin by removing some columns of summary statistics that we wish to ignore. Selecting a list of column names with a "minus" - sign in front of the list will remove these items from the dataframe and keep the rest in place. The command below pipes the modified dataframe to the `kable` command in the *knitr* package for a more pleasing tabular display.

```{r}
# Let's observe the contents of profiling_num:
funModeling::profiling_num(gapminder) %>% 
  dplyr::glimpse()
```

```{r}
# now remove unwanted columns from summary display
funModeling::profiling_num(gapminder) %>%
  select(-c("variation_coef","skewness","kurtosis","range_98","range_80","p_01","p_99")) %>%
  knitr::kable()
```

```{r}
funModeling::profiling_num(gapminder) %>%
  select(c("variable","mean","std_dev","p_25","p_50","p_75")) %>%
  knitr::kable()
```

## 4.6 Order using arrange

```{r}
# This command will show the countries with highest life expectancy because 
# the data are arranged in descending order of life expectancy (larger to smaller)
gapminder %>% 
  dplyr::filter(year==1997) %>%   
  dplyr::select(country, continent, lifeExp) %>% 
  dplyr::arrange(desc(lifeExp)) %>% 
  head()
```

```{r}
# This command uses the default ascending (increasing) order with
# respect to life expectancy (order smaller to larger)
gapminder %>% 
  filter(year==1997) %>%  
  select(country, continent, lifeExp) %>% 
  arrange(lifeExp) %>% 
  head()
```

## 4.7 Grouped Filter

Another useful verb in the *tidyverse* is `group_by`. Suppose we wanted to view the two countries with the highest life expectancy in 1997, in each continent.

```{r}
gapminder %>% 
  filter(year==1997) %>%  
  select(country, continent, lifeExp, pop) %>%
  dplyr::group_by(continent) %>%
  dplyr::top_n(n=2,wt=lifeExp) %>%
  dplyr::arrange(continent) %>%
  knitr::kable()
```

## 4.8 New Variables Using Mutate

```{r}
gapminder %>%
  dplyr::mutate(logpopulation = log(pop)) %>%
  dplyr::glimpse()
```

```{r}
gapVers1 <- gapminder %>%
  dplyr::mutate(logpopulation = log(pop)) %>%
  dplyr::rename(logPop=logpopulation) 
#
  dplyr::glimpse(gapVers1)
```

The `if_else` function from *dplyr* has the form \`if_else(logical condition,value if TRUE, value if FALSE).

The next example uses the "T-pipe" function `%T>%` to break the piping so that the result of the second `mutate` flows to both `glimpse` and to `head` - in this construction, it is understood the output of `glimpse` does not pipe to `head`, but rather the original data flow from the second `mutate`.

```{r}
gapminder %>% 
  dplyr::mutate(region = if_else(continent=="Oceania","Oceania","NotOceania")) %>%
  dplyr::mutate(regionf = as_factor(region)) %T>%
  dplyr::glimpse() %>%
  head()
```

## 4.9 Simple Counting Using tally() and count()

We frequently wish to know how many observations/rows satisfy a set of conditions. We will filter the observations for the given conditions, then count them using the `tally()` or `count()` functions from *dplyr*.

Essentially, `count()` is a short-hand for `group_by() + tally()`.

For example, what if we want to know how many observations are from continent 'Americas' in 1997.

These examples have no grouping, no `group_by` is being used.

```{r}
gapminder %>% dplyr::filter(year==1997) %>%
  dplyr::filter(continent=="Americas") %>%
  dplyr::tally()

```

```{r}
gapminder %>% dplyr::filter(year==1997) %>%
  dplyr::group_by(continent) %>%
  dplyr::filter(continent=="Americas") %>%
  dplyr::tally()
```

## 4.10 Missing Values

```{r}
x <- c(1,2,NA,4)
y <- c(11,12,13,NA)
z <- c(7,8,9,10)
tempdf <- data.frame(x,y,z)
tempdf

```

```{r}
# count missing values for variable x
tempdf %>%
  dplyr::summarise(count = sum(is.na(x)))
```

```{r}
# count rows with missing y
tempdf %>%
  dplyr::tally(is.na(y))
```

```{r}
# subset of rows with complete data for specified columns
tempdf %>%
  dplyr::select(y,z) %>%
  tidyr::drop_na() %>%
  head()
```
