---
title: "EDA_bank2"
author: "William HU ZIHAO"
date: "2024-10-08"
output:
  pdf_document: default
---

## Step 1 Load Dataset

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
library(conflicted)
library(tidyverse)
library(lubridate)
library(bizdays)
conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
knitr::opts_chunk$set(echo = TRUE, fig.align = "center", collapse = TRUE, cache = TRUE)
bank <- read_delim("https://bit.ly/4d8jY4I", ";")
bank <- bank %>%
  mutate(y = factor(y, levels = c("no", "yes")))
```

## Step 2 Exploratory Data Analysis

### 2.1 Fundamental Analysis

```{r}
View(bank)
# Preview the structure of the data
glimpse(bank)
```

```{r}
# Show summary statistics of the data
summary(bank)
```

### 2.2 Distribution of Categorical Variables

```{r}
bank %>%
  select_if(is.character) %>%
  map(~table(.) %>% as.data.frame()) %>%
  imap(~ setNames(.x, c(.y, "Count")))

```

### 2.3 Distribution of Numerical Variables

```{r}
bank %>%
  select_if(is.numeric) %>%
  gather(key = "variable", value = "value") %>%
  ggplot(aes(x = value)) +
  geom_histogram(bins = 30, fill = "blue", color = "black", alpha = 0.7) +
  facet_wrap(~variable, scales = "free") +
  theme_minimal() +
  labs(title = "Distribution of Numeric Variables")

```

### 2.4 Balance vs. Subscription

```{r}
# Plot balance distribution using boxplot
bank %>%
  mutate(y = factor(y, levels = c("no", "yes"), labels = c("No Subscription", "Subscribed")))%>% ggplot(aes(x = y, y = balance, fill = y)) +
  geom_boxplot() +
  labs(title = "Balance Distribution by Term Deposit Subscription",
       x = "Term Deposit Subscription",
       y = "Average Yearly Balance (in euros)") +
  theme_minimal() +
  scale_fill_manual(values = c("No Subscription" = "tomato", "Subscribed" = "steelblue")) +
  theme(legend.position = "none")

```

Notice that the visulizations are greatly influenced by the outliers. Thus, we can consider removing outliers in two subscription groups for a more direct visulizations of balance versus two subscription groups.

```{r}
remove_outliers <- function(data) {
  Q1 <- quantile(data$balance, 0.25, na.rm = TRUE)
  Q3 <- quantile(data$balance, 0.75, na.rm = TRUE)
  IQR <- Q3 - Q1
  lower_bound <- Q1 - 1.5 * IQR
  upper_bound <- Q3 + 1.5 * IQR
  data %>%
    filter(balance >= lower_bound & balance <= upper_bound)
}

bank_filtered <- bank %>%
  mutate(y = factor(y, levels = c("no", "yes"), labels = c("No Subscription", "Subscribed"))) %>%
  group_by(y) %>%
  group_modify(~ remove_outliers(.x)) %>%
  ungroup()

ggplot(bank_filtered, aes(x = y, y = balance, fill = y)) +
  geom_boxplot() +
  labs(title = "Balance Distribution by Term Deposit Subscription (Without Outliers)",
       x = "Term Deposit Subscription",
       y = "Average Yearly Balance (in euros)") +
  theme_minimal() +
  scale_fill_manual(values = c("No Subscription" = "tomato", "Subscribed" = "steelblue")) +
  theme(legend.position = "none")
```

### 2.5 Density Plot of Contact Duration by subscription behavior of different job types

```{r}
# Step 1: Calculate the median duration for 'yes' subscription by job
medians_sub <- bank %>%      
  filter(y == "yes") %>%      
  group_by(job) %>%      
  summarize(median_duration = median(duration, na.rm = TRUE), .groups = 'drop')

# Step 2: Order jobs based on median duration for 'yes' subscription
ordered_jobs <- medians_sub$job[order(medians_sub$median_duration)]

ggplot(bank, aes(x = duration, fill = y)) +   
  geom_density(alpha = 0.7) +     
  facet_wrap(~factor(job,levels = ordered_jobs), scales = "free_y") +   
  labs(title = "Contact Duration Distribution by Job Type and Subscription Status (ordered by median duration of subscirbers)",        x = "Duration (seconds)",        y = "Density",        fill = "Subscription Status") +   
  theme_minimal()  
```

### 2.6 Density Plot of Duration of Subscribers by Month

```{r}
subscriptions <- bank %>%
    filter(y == "yes") %>%
    mutate(month = factor(month, 
                          levels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")))  # Ensure month is ordered


# Create the density plot
ggplot(subscriptions, aes(x = duration, fill = month)) +
    geom_density(alpha = 0.5) +  # Density plot with transparency
    labs(title = "Density of Subscription Durations by Month",
         x = "Duration (seconds)",
         y = "Density",
         fill = "Month") +
    theme_minimal() +
    theme(legend.position = "right") +
    scale_fill_brewer(palette = "Set3")
```

```{r}
subscriptions <- bank %>% 
  filter(y == "yes") %>%
  mutate(month = factor(month, 
                        levels = c("jan", "feb", "mar", "apr", "may", "jun", "jul", "aug", "sep", "oct", "nov", "dec")),
         quarter = case_when(
           month %in% c("jan", "feb", "mar") ~ "Q1 (Jan-Mar)",
           month %in% c("apr", "may", "jun") ~ "Q2 (Apr-Jun)",
           month %in% c("jul", "aug", "sep") ~ "Q3 (Jul-Sep)",
           month %in% c("oct", "nov", "dec") ~ "Q4 (Oct-Dec)"
         ))

# Create the density plot with facets for each quarter
ggplot(subscriptions, aes(x = duration, fill = month)) +
  geom_density(alpha = 0.5) +  
  labs(title = "Density of Subscription Durations by Quarter",
       x = "Duration (seconds)",
       y = "Density",
       fill = "Month") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3") +
  facet_wrap(~quarter)
```

### 2.7  (No) Subscriptions by Contact and Age

```{r}
library(patchwork)
sub_percentage_contact<-bank %>%
  filter(y == "yes") %>%
  group_by(age, contact) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(age) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()%>%
# Create the stacked bar plot (3D-like view) with X-axis as percentage, Y-axis as age, and fill by contact type
ggplot(aes(x = age, y = percentage, fill = contact)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(title = "Subscription Proportions by Contact Type and Age",
       x = "Age",
       y = "Percentage of Subscriptions",
       fill = "Contact Type") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")  


no_sub_percentage_contact<-bank %>%
  filter(y == "no") %>%
  group_by(age, contact) %>%
  summarise(count = n()) %>%
  ungroup() %>%
  group_by(age) %>%
  mutate(percentage = count / sum(count)) %>%
  ungroup()%>%
# Create the stacked bar plot (3D-like view) with X-axis as percentage, Y-axis as age, and fill by contact type
ggplot(aes(x = age, y = percentage, fill = contact)) +
  geom_bar(stat = "identity", position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) + 
  labs(title = "No Subscription Proportions by Contact Type and Age",
       x = "Age",
       y = "Percentage of No Subscriptions",
       fill = "Contact Type") +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")  

sub_percentage_contact+no_sub_percentage_contact
```

## Step 3 Conclusions

### 3.1 Average Yearly Balance by Subscription Group

*Code & Results are in* [2.4 Balance vs. Subscription]

-   **Balance Discrepancy:** Clients who subscribed to a term deposit generally have a higher balance compared to those who did not subscribe. The wider IQR in the subscribed group indicates more variability in the financial status of these clients.

-   **Outliers**: While there are extreme cases (outliers) in both groups, the extreme balance values of non-subscribers' group which stretch beyond 20000 euros indicate that while the majority of non-subscribers have low balances, a small number of them have much higher balances than subscribers.

-   **Non-subscribers**

    -   **Median Balance**: The median is slightly above zero, indicating that for most clients who did not subscribe, their average balance is quite low.

    -   **Spread of Balance**: The interquartile range (IQR) is narrow, suggesting that most clients in this group have balances concentrated around 0-1000 euros.

-   **Subscribers**

    -   **Median Balance**: The median is higher than that of the non-subscribers, around 600 euros, suggesting that clients who subscribed tend to have higher balances.

    -   **Spread of Balance**: The IQR is wider compared to the non-subscribers, indicating a more varied distribution of balances in this group.

### 3.2 Contact Contact Duration by subscription behavior of different job types {#proportion-of-housing-personal-loan-by-martial-group}

*Code & Results are in* [2.5 Density Plot of Contact Duration by subscription behavior of different job types]

-   **General Observation**: Across most job types, longer contact durations are associated with a higher likelihood of subscription to the term deposit. This is evident by the increased density of subscribers (in teal) for longer durations, whereas non-subscribers (in red) peak at shorter durations.

-   **Job-specific Insights**:

    -   **Admin, Blue-collar, Entrepreneur, Housemaid, and Services**: These jobs show a strong trend where short contact durations lead to more non-subscriptions. The density of subscribers rises significantly as the duration increases.
    -   **Management, Retired, Self-employed, and Unemployed**: Similarly, there is a noticeable increase in subscription rates as contact duration increases. However, these groups also show some variation, with more balanced densities at certain durations.

-   **Short Durations**: For nearly all job types, very short contact durations (under 100 seconds) are predominantly associated with non-subscription.

### 3.3 Duration of Subscribers by Month

*Code & Results are in* [2.6 Density Plot of Duration of Subscribers by Month]

-   **Seasonal Trend of Engagement of Subscribers:**

    -   **High Engagement during Q2 and Q3**

        -   From density plot of duration of subscribers by quarter, it can be concluded that the campaigns usually experience **high engagements during Q2 and Q3** which could be reflected by higher median and mean of contact duration

        -   Understanding that Q2 and Q3 lead to longer subscription durations can guide marketing efforts. Resources can be allocated to intensify campaigns during those times to maximize subscriptions.

    -   **Low Engagement during Q1 and Q4**

        -   **Q1**: The lower durations in Q1 could suggest that new year resolutions or fresh starts motivate customers to subscribe, but their engagement might not be sustained, leading to shorter subscription durations.

        -   **Q4**: In Q4, the lower durations could be influenced by year-end behaviors, holiday distractions, or other seasonal factors that might lead to shorter subscription periods as customers may not prioritize subscription activities during festive seasons.

### 3.4 (No) Subscriptions by Contact and Age

*Code & Results are in* [2.7 (No) Subscriptions by Contact and Age]

#### **Subscription Proportions by Contact Type and Age**

-   **Cellular (teal)**:

    -   Cellular subscriptions dominate across all age groups, especially for younger people (under 50). The proportion remains high even in older age groups, though it slightly decreases as age increases.

-   **Telephone (yellow)**:

    -   Telephone subscriptions become more prominent in people aged 50 and above, especially among older individuals, indicating that traditional landlines are more commonly used by older populations.

-   **Unknown (purple)**:

    -   The "unknown" category shows a slight increase in the middle-aged to older groups (around 50 to 70). However, its overall proportion is low compared to cellular and telephone subscriptions.

#### **No Subscription Proportions by Contact Type and Age**

-   **Cellular (teal)**:

    -   The "No Subscription" plot shows that even among those without subscriptions, cellular contact information remains the most common contact type across all age ranges.

-   **Telephone (yellow)**:

    -   Landline (telephone) contacts without subscriptions become increasingly prominent as age increases, particularly in those over 50, with a slight peak at around 70 years.

-   **Unknown (purple)**:

    -   The proportion of "unknown" contact types rises significantly for people without subscriptions in middle-aged (40 to 60) and older groups, but it's relatively rare in younger age groups.
